---
title: "ConnTextUL"
output: html_document
date: "2023-06-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
requires('scripts/requirements.txt')
source('scripts/utilities.R')
source('scripts/load.R')

```

# Phonological predictions (Variable: `phon_prediction_features`)
The phonological predictions are written in the file as strings with phonemes separated by `;`. In order to parse those we need to separate the phonemes based on the presence of `;` and convert to numeric. We process those into (at least) two different forms: one version with all phonemes concatenated into a single numeric vector, and a second where the phonemes are preserved as separate segments. Note that the predictions are not padded, so the prediction for a word with 5 phonemes will be different than that with 7 phonemes, where the prediction (of course) should match the length of the target.

```{r}
View(head(preds))

```

```{r phoneme_features}
phon_features = c("labial", "dental", "alveolar", "palatal", "velar", "glottal", "stop", "fricative", 
             "affricate", "nasal", "liquid", "glide", "rhotic", "tap", "voice", "front", "center",
             "back", "close", "close-mid", "mid", "open-mid", "near-open", "open", "tense", "retroflex",
             "round", "post_y", "post_w", "primary", "secondary", "sos", "eos")

```

# Phonological prediction probabilities (Variable: `phon_prediction_probabilities`)



# Analyses by phoneme
Here is the tabulation of mistakes over phonemes.
```{r}
preds_by_phoneme %>% 
  filter(correct == 0,
         condition == "predicted") %>% 
  select(all_of(phon_features)) %>% 
  pivot_longer(everything()) %>% 
  ggplot(aes(name, value)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
